---
title: 'R basics: data and functions'
author: "Kasper Welbers, Wouter van Atteveldt & Nel Ruigrok"
date: "November 2023"
output: 
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
urlcolor: blue
---


```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(echo = TRUE, results = FALSE, message = FALSE, warning = FALSE)
#library(printr)

```


# This tutorial

Data science is not always smooth sailing. You will often find yourself willing to do something specific that is not directly supported by a package, or you have to work with data that first needs to be cleaned and prepared in order to use a certain function. It is therefore important to learn some of the basics. 

In this week's lab you will learn about basic R data types, data structures and functions. You can consider these as the main building blocks for working with R data. Learning about these basics might be less exciting compared to using the advanced features, but in the long run it will save you time and frustration. 
An additional R Markdown file is available in the course material for practicing with data types, data structures and functions.

# Data types

Data types concern the different types that single values in data can have. The most basic data types in R are: 

* numeric (numbers)
* character (text)
* factor (categorical data)
* logical (True or False)

In addition, there are special types for data such as date/time values.

* Date (calendar dates) or POSIXlt (calendar dates and times). 

## Numeric

Numbers. As simple as is gets. You can use them to do the math you know and love.

```{r}
x = 5      ## assign a number to the name x
class(x)   ## view the class of the value assigned to x

x + 3
x / 2
log(x)     ## logarithm
sqrt(x)    ## square root
```

For those who have experience with low-level programming languages, it is nice to know that you do not need to think about different types for representing numbers (int, double, float, etc.). 

## Character

Textual data, either as single characters, entire words, or even full texts.

```{r}
x = "Some text"  ## assign text to the name x
class(x)         ## view the class of the value assigned to x
```

It's important to recognize the distinction between names and character values. In the above example, x is the name to which the text "some text" has been assigned. Whether a word is a name or a character value is indicated with quotes.

```{r}
x       ## get value assigned to the name x
"x"     ## the text "x"
```

Naturally, you cannot perform math with character data. 
Using the wrong data type will generally yield an error, as seen here.

```{r, eval=F}
sum(x)
```

It's import to recognize these types of errors, because they are terribly common.
You might have imported data in which a column that's supposed to contain numbers accidentally contains a word,
in which case R will consider the column to be column of character values. 

Note that you can express a number as a character value, e.g., "1", "999", but not a text as a numerical value. If it is possible to convert a value to a different type, you can do so with the **as** method:

```{r, eval=F}
x = "999"
x = as.numeric(x)     ## converts character to numeric
x

y = 999
y = as.character(y)   ## converts numeric to character
y

z = "nein nein nein"
z = as.numeric(z)     ## tries to convert character to numeric, but fails 
z
```

R has decent built-in support for working with character values, but for more advanced techniques for working with **strings** it is recommended to use a dedicated package such as [stringr](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf).

## Additional types

Additional instructions for working with factors, logical and date type values are included at the bottom of this document. But for now we'll first move on to data structures.

# Data structures

In SPSS or Excel, data is always organized in a rectancular data frame, with cells arranged in rows and columns.
Typically, the rows then represent cases (e.g., repondents, participants, newspaper articles) and columns represent variables (e.g., age, gender, date, medium).
For most analyses, this is also the recommended data format in R, using the *data.frame* structure. 
However, an important difference is that in R it is possible, and often usefull, to combine different formats.
Also, to understand how a *data.frame* in R works, it is usefull to understand that a *data.frame* is a collection of *vectors*, and thus it is usefull to understand how *vectors* work.  

Here we will first briefly discuss *vectors*, and then quickly move on to *data.frames*. In addition, we will mention *matrices* and *lists* on a good-to-know-about basis.


## Vector

The concept of a vector might be confusing from a social science background, because we rarely use the term in the context of statistics (well, not consciously at least). We won't address why R calls them vectors and how this relates to vector algebra, but only how you most commonly use them.

A vector in R is a **sequence** of **one or more values** of the **same data type**
From a social science background, it is very similar to what we often call a **variable**. 

You can declare a vector in R with c(...), where between the parentheses you enter the elements, separated with commas. 
The number of elements is called the length of the vector.
A vector can have any of the data types discussed above (numeric, character, factor, logical, Date).

```{r}
v1 = c(1, 2, 10, 15)    ## a numeric vector of length 4
v2 = c("a", "b", "b")   ## a character vector of length 3
v3 = 1:10               ## a numeric vector of length 10 with the values 1 to 10. 
```

If you combine data types in the same vector, R will generally use the broadest data type for the entire vector. For example, we saw earlier that a number can be expressed as a character value, but a text cannot be expressed as a numerical. Accordingly, if we combine both types in a vector, R will convert the numerical values to character values.

```{r}
c(1, 2, "c")            ## becomes a character vector of length 3
```

Since vectors can only have one type, we can perform type specific operations with them. In many ways, we can work with them in the same way as we can work with single values. In fact, single values are actually just vectors of length 1. For example, if we have a vector of numeric type, also called a numeric vector, we can perform calculations.

```{r}
x = c( 1, 2, 3, 4, 5)
y = c(10,20,30,40,50)
x + y     ## for 2 vectors of same size calculations are pairwise (1 + 10, 2 + 20, etc.)
x + 10    ## for a vector and single value, the value is repeated (1 + 10, 2 + 10, etc.)
```


## Data.frame

A *data.frame* is essentially a collection of vectors with the same length, tied together as columns. To create the *data.frame*, we use the **data.frame()** function. We enter the vectors in the format: `column_name = vector`. Here we create a data.frame for data from a fictional experiment.

```{r, results=T}
d = data.frame(id =        1:10,
               condition = c('E', 'E', 'C', 'C', 'C', 'E', 'E', 'E', 'C', 'C'),
               gender =    c('M', 'M', 'F', 'M', 'F', 'F', 'F', 'M', 'M', 'F'),
               age =       c( 17,  19,  22,  18,  16,  21,  18,  17,  26,  18),
               score_t1 =  c(8.0, 6.0, 7.5, 6.8, 8.0, 6.4, 6.0, 3.2, 7.3, 6.8),
               score_t2 =  c(8.3, 6.4, 7.7, 6.3, 7.5, 6.4, 6.2, 3.6, 7.0, 6.5))
d
```

Now, the data structure clearly implies that there is a relation between the elements in the *column vectors*.
In other words, that each row represents a *case*. In our example, these cases are participants, and the columns represent:

* the participant **id**.
* the experimental **condition** (E = experimental condition, C = control group) 
* demographic variables: **gender** and **age**.
* test scores before and after the experimental condition: **score_t1** and **score_t2**


# Functions

Where data types and structures concern how data is respresented in R, functions are the tools that you use to read, create, manage, manipulate, analyze and visualize data. 

## What is a function?

There are many correct and formal ways to define what functions are, but for the sake of simplicity we will focus on an informal description of how you can think of functions in R:

* A function has the form: `output = function_name(argument1, argument2, ...)`
    + **function_name** is a name to indicate which function you want to use. It is followed by parentheses.
    + **arguments** are the input of the function, and are inserted within the parentheses. Arguments can be any R object, such as numbers, strings, vectors and data.frames. Multiple arguments can be given, separated by commas. 
    + **output** is anything that is returned by the function, such as vectors, data.frames or the results of a statistical analysis. Some functions do not have output, but produce a visualization or write data to disk.

* The purpose of a function is to make it easy to perform a (large) set of (complex) operations. This is crucial, because
    + It makes code easier to understand. You don't need to see the operations, just the name of the function that performs them
    + You don't need to understand the operations, just how to use the function
 
For example, say that you need to calculate the square root of a number. This is a very common thing to do in statistical analysis, but it actually requires a quite complicated set of operations to perform. This is when you want to use a function, in this case the `sqrt` (square root) function. 

```{r}
sqrt(5)
```

In this example, the function name is `sqrt`. The input is the single argument `5`. If you run this code, it produces the output `2.236068`. Currently, R will simply print this output in your Console, but as you learned before, we can assign this output to a name.

```{r}
square_root = sqrt(5)
```

This simple process of input -> function -> output is essentially how you work with R most of the times. You have data in some form. You provide this data as input to a function, and R generates output. You can assign the output to a name to use it in the next steps, or the output is a table with results or a visualization that you want to interpret.  

## Using functions

Above you saw the simple function `sqrt()`, that given a single number as input returned a single number as output. As you have also seen in the first week, functions can have multiple arguments as input. Recall the following function from the `quanteda` package. You don't have to run the code this time, just try to recognize the arguments. 

```{r, eval=F}
dfm(x = "some text", tolower = TRUE, stem=TRUE)
```

This function, with the name `dfm`, is given several arguments here: `x`, `tolower` and `stem`. Given this input, many operations are performed behind the scenes to create a document-term matrix.

By now we hope you have realized just how broad the use of functions is. The *R syntax* for performing basic mathematical operations such as `sqrt()` is essentially the same as the syntax for creating a document-term matrix, performing advances statistical analysis or creating powerfull visualizations. Accordingly, if you understand this syntax, you can do almost anything in R. 

The many R packages that you can install are mostly just collections of functions (some also provide new **classes**, which we'll save for later). We will now show how you can learn how to use each function by knowing how to view and interpret it's documentation page.


### Viewing and interpreting function documentation

You can access the documentation of a function by typing a question mark in front of the function name, and running the line of code. Let's do this to view the documentation of the `sqrt()` function

```{r}
?sqrt
```

If you run this in RStudio, the help page will pop-up in the bottom-right corner, under the *Help* tab page. Sometimes, if the name of a documentation page is used in multiple packages, you will first receive a list of these packages from which you will have to select the page. 

For the `sqrt()` function, the help page has the **title** "Miscellaneous Mathematical Functions".
Just below the title, you see the **Description**, in which the author of a function briefly describes what the function is for. 
Here we see that there are two functions that are grouped under "Miscellaneous Mathematical Functions", the `abs()` function for computing the absolute value of a number `x`, and the `sqrt()` function for the square root.  

Under description, the **Usage** is shown. This is simply the name of the function or functions, and the possible arguments that you can use. Here the Usage is extremely simple: both functions only take one argument named `x`. In a minute, we'll discuss functions with multiple arguments.

Below usage, the **Arguments** section explains how to use each argument. Here, the only argument is `x`, and it is explained that x is "a numeric or complex vector or array". For now, let's focus only on the case of a numeric vector. It appears that in addition to giving a single value like above (recall that in R this is actually a vector of length 1) we can give a vector with multiple numbers.

```{r}
sqrt(c(1,2,3,4,5))
```

There are more parts to the documentation that we'll ignore for now. Notable parts to look into for yourself are **Details**, that provides more information, and the **Examples** section at the very bottom, which is a great starting point to see a function in action. 



# Additional instructions

Here we provide several additional instructions  about data types, structures and functions. There are all very usefull to know about, but a bit more hard to chew. Prioritize understanding the former part, but by all means, do study these a additional instructions.

## Additional data types

### Factor

A *factor* in R is a series of labeled numbers.
This is usefull if you have a categorical variable, such as education level (in surveys) or medium type (in content analysis).
If you are familiar with SPSS, this is comparable to using *value labels*. 

```{r}
x = c('De Volkskrant','De Volkskrant','NRC Handelsblad',
      'De Volkskrant','NRC Handelsblad','Trouw')
x
```

We now have a sequence of character values. What's special about these character values, is that the same values are often repeated. In this case, it's best to think as each unique value ("De Volkskrant", "NRC Handelsblad", "Trouw") as a label. This is a typical job for the *factor* type, so let's transform the character type vector to a factor type.

```{r}
x = factor(x)
x
```

Two things have changed. 
Firstly, you see that there is now a line that says "levels: ...", in which the three unique labels are shown.
Secondly, the quotes have disappeared. This is because it's no longer a character value.

Behind the scenes, x is now a sequence of numbers, where each number points to a label

```{r}
as.numeric(x) ## show the numbers: 1 for De Volkskrant, 2 for NRC and 3 for Trouw
levels(x)     ## show the levels / labels. 
```

If this confuses you, you're still perfectly healthy. The benefits of factors become more apparent later on when you start working with certain types of analysis, visualizations, and when you use very large data (numeric values require less memory than character values). 

You might have concluded that factors are only a hassle, and you'll simply stick to character values.
This is a valid strategy, except that R tends to force factors on you, for example when you import data or make a data.frame, and R thinks that your character column is better of as a factor. There are ways to ask R not to do this, but really, you're better of just accepting factors. It's better on the long run.

Still, if you ever run into trouble with factors and really prefer to use character values, you can simply convert them into "character" type, using **as.character()**

```{r}
as.character(x)
```

### Logical

The *logical* data type only has two values: **TRUE** and **FALSE** (which can be abbreviated to **T** and **F**). You will not often encounter logical values in your data, but you will use them in many operations such as subsetting and transforming data, as you will see when we discuss data structures. Understanding a bit about logical values will help you understand how these operations work.

For now, we will focus on how logical values result from comparisons, and how logical operators can be used. 

### Date

Date is not one of the basic data types in R, but in social scientific research we often work with calendar dates and times. This requires a special data type, because there are many limitations or complications if we try to express dates and times as character or numeric values. Two of the most common date classes in native R (i.e. not requiring additional packages) are *Date*, which only handles calendar dates, and *POSIXlt*, that handles both calendar dates and times. 

In this tutorial we will only demonstrate how to use *POSIXlt*, but the general approach for working with dates and times in R is the same for different date types (and is actually similar in other programming languages as well). You can do most things that you need with only two functions: *strptime* and *strftime*

* strptime (**str**ing **p**arse **time**) creates a date/time value from a character value
* strftime (**str**ing **f**ormat **time**) extracts parts of a date/time from a date/time value.

For both functions, you need to know how to specify the *format*. Essentially, this is a string (character value) that contains the date format, using special placeholders to indicate specific parts of the date. These placeholders are always a percentage symbol **%** followed by a letter. The most commonly used placeholders are:

placeholder          | date part
-------------------- | ------------------------
%Y | year with century (2010, 2011, etc.)
%m | month as a decimal number (01, 02, 03, ..., 10, 11, 12)
%d | day as a decimal number (01, 02, 03, ..., 28, 29, 30)
%H | hour as a decimal (00, 01, 02, ..., 21, 22, 23)
%M | minute as a decimal (00, 01, 02, ..., 57, 58, 59)
%S | seconds as a decimal (00, 01, 02, ..., 57, 58, 59)

Using these placeholders, we can describe various date formats. Here we show how to use this together with the **strptime** function to **parse** a date from character type to a date (POSIXlt) type. We will explain more about functions below. For now, just note that the first argument passed to **strptime()** is the date, and the second argument is the format.

```{r}
strptime('2010-01-01 20:00:00', format = '%Y-%m-%d %H:%M:%S')
strptime('01/01/2010', format = '%m/%d/%Y')
strptime('2010 any 01 format 01 goes', format = '%Y any %m format %d goes')
```

With the **strftime()** function we can use the same format strings to extract specific parts from a date/time value. Here, we first create a POSIXlt date with strptime, and then use strftime to extract parts. Note that all parts are returned as a character value, even if they are single numbers.

```{r}
x = strptime('2010-06-01 20:00:00', format = '%Y-%m-%d %H:%M:%S')
strftime(x, '%Y')
strftime(x, '%Y week %W')
strftime(x, 'Today is a %A')  ## language of weekday depends on your locale settings
```

You can find a full list of format placeholders in the documentation of the **strptime** function, which you can access by putting a questionmark in front of the function name and running it as code. The help page should open in your bottom-right window in RStudio.

```{r, eval=F}
?strptime
```

For reference, if you feel the urge to master date and time operations, there is also the excellent **lubridate** package. Usage is explained in detail in this (free online) [chapter from "R for Data Science"](http://r4ds.had.co.nz/dates-and-times.html). You can also download the cheatsheet [here](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf).

## Comparisons

The most common way in which you will get logical values in R is as the results of a **comparison** between two values or two vectors. There are 6 types of comparissons, which use operators that you are probably already familiar with:

operator   | meaning
---------- | --------------------
x < y      | x is smaller than y             
x > y      | x is larger than y               
x <= y     | x is smaller than or equal to y  
x >= y     | x is larger than or equal to y     
x == y     | x equals y                       
x != y     | x does not equal y               

The outcome of a comparison of two values is a logical value. 

```{r}
5 < 10
5 < 2
```

The outcome of a comparison of two vectors is a logical vector. For example, let's assume that we have the average math grade's for 5 students in year 1 and year 2, and we look-up whether their grade went down (i.e. grade in year 1 was higher than year 2). Here we see that only for the first student the grade went down (from 6 in year 1 to 5 in year 2), so only the first value is TRUE. 

```{r}
grade_year_1 = c(6,6,7,7,4)
grade_year_2 = c(5,6,8,7,6)
grade_year_1 > grade_year_2    
```

If we compare a vector to a single value, each value of the vector will be compared to that value. As you will see shortly, this is a core mechanic behind selecting and subsetting data. In the following example, we will look-up which participants for our hypothetical drinking study are at least 18 years old.

```{r}
age = c(17,34,12,20,12,25,30,14,33)
age >= 18
```

Finally, while not strictly a comparison, there is another common operator that creates a logical value or vector.

operator   | meaning
---------- | --------------------
x %in% y   | the value(s) in x also exist in y             

This is for instance usefull if you want to look for multiple values. For example, let's select select `Bob` and `Carol` from our list of names.

```{r}
name = c('Alice','Bob','Carol','Dave','Eve')
name %in% c('Bob', 'Carol')
```


## Logical operators

Given a logical value (or vector), we can use the logic operators & (AND), | (OR) and ! (NOT). In the following table, x and y are both logical values.

operator   | meaning | outcome
---------- | ------- | --------------------
x & y      | x AND y | only TRUE if both x and y are TRUE           
x \| y     | x OR y  | only TRUE if either x or y, or both are TRUE.               
!x         | NOT x   | the opposite if x, so only TRUE if x is FALSE.

You can play with the following code to try it out (results not shown in this document).

```{r, eval=F}
TRUE & TRUE        ## is TRUE
TRUE & FALSE       ## is FALSE
FALSE & FALSE      ## is FALSE

TRUE | TRUE        ## is TRUE
TRUE | FALSE       ## is TRUE
FALSE | FALSE      ## is FALSE

!TRUE              ## NOT TRUE, so is FALSE
!FALSE             ## NOT FALSE, so is TRUE

TRUE & !FALSE      ## !FALSE == TRUE, so you get TRUE & TRUE
TRUE & (!TRUE | TRUE) ## !TRUE | TRUE == FALSE  | TRUE == TRUE, so you get TRUE & TRUE 

a = c(T, T, F, F, F, T)
b = c(T, F, F, F, T, T) 
a & b              ## 6 pairwise comparisons: T & T, T & F, F & F, F & F, F & T, T & T) 
```


## Methods, generic functions and the three dot ellipsis

### A note about 'methods' and 'generic functions'

Some functions are *generic functions*, that use different *methods* depending on the input that they are used with. Ignoring technicalities, there's one thing you currently need to know about them, because you will need it to interpret their documentation pages. 

A method is a function that is associated with a specific object. For example, subsetting a `vector` works differently from subsetting a `data.frame` or `matrix`. Still, it is convenient to only have one function called `subset()` that can be used on all these kinds of input. In R, the `subset()` functions is therefore a *generic function*, that will behave differently depending on the kind of input. 

The type of input to the `subset()` function therefore also determines what type of arguments can be used. You can see this in the documentation page.

```{r}
?subset
```

In the description we see that `subset()` can be used on vectors, matrices or data.frames. The **Usage** section therefore contains different versions, for different *S3 methods* (ignore "S3" for now) that are associated with different kinds of input. The general form is `subset(x, ...)`, which shows that subset always requires an argument `x`, and in the **Arguments** we see that `x` is the "object to be subsetted". We then see three methods: default, 'matrix' and 'data.frame'.

* The default will be used if `x` is neither a `matrix` or `data.frame` (for instance a `vector`). In this case the only argument is *subset*, which is the expression (e.g., `x > 10`) used to make a selection. 
* If the input is a 'matrix', there are two additional arguments: *select* and *drop*. It makes sense that these are not available for vectors, because they are both only relevant if there are multiple columns. That is, *select* is used for selecting columns, and *drop* can be used to have subset return a `vector` (instead of a `matrix`) if only one row or column remains after subsetting.
* If the input is a 'data.frame', the same arguments are used as for 'matrix' (but internally the method works differently)

### The special case of the three dot ellipsis

A special type of argument that you'll often encounter in function documentation is the three dot ellipsis (`...`). This is used to pass any number of named or unnamed arguments. A good example of how this is used, is in the `data.frame()` function. In a previous tutorial you saw that you can use this function to create a data.frame from vectors, where names are used as column names. Now, you will see that these are actually just *named arguments*.

```{r}
?data.frame
data.frame(x = 1:5, y = c('a','b','c','d','e'))
```

As an additional example, consider the `sum()` function. 
Here the `...` is used for "numeric or complex or logical vector". This means that we can add any number of arguments with numbers in them, and they will all be added up.

```{r}
?sum
sum(1, 2, 3, c(1,2,3))
```

To clarify, if we want to set any of the other arguments, such as `na.rm` in `sum()`, we can do so by referring to them by name. By default, `sum()` returns NA (R's way of saying "missing") if any NA is present. As noted in the documentation, we can ignore the NA values by setting `na.rm` to `TRUE`

```{r}
sum(1,2,NA)
sum(1,2,NA, na.rm = T)
```

Finally, a way in which the three dot ellipsis is also often used, is to pass arguments on to another function that is used within the function. If you look back at the documentation for the `dfm()` function, you'll see in the explanation of `...`: "additional arguments passed to tokens; not used when x is a dfm". In this case, you can see which arguments these are by looking at the documentation of the `tokens` function. Here you see that you could also pass the argument `remove_numbers = TRUE` to `dfm()`.


# Further reading

If you want to learn more about the basics of R, we recommend:

* [The (free online) R for Data Science book](http://r4ds.had.co.nz/functions.html)
* Taking one of the many (free) online introductions, such as the ones offered at [DataCamp](https://www.datacamp.com/courses/free-introduction-to-r/) or [Code School](http://tryr.codeschool.com/). 

Also, it could be nice to grab some of the [cheatsheets](https://www.rstudio.com/resources/cheatsheets/) collected on the RStudio website. For the basics of R, the cheatsheet [Base R](http://github.com/rstudio/cheatsheets/raw/master/base-r.pdf) is particularly usefull.